Pyrovelocity PyPI user notebook template

### 2024-03-21 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
fix(pypi_template): add exploration section
- temporarily install from github
- use internal help
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

Installation should take less than **5 minutes**. 
It involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
Otherwise, it is assumed that the user will have installed pyrovelocity in the kernel environment.

@@

### Install pyrovelocity

@@

def setup_pyrovelocity():
    if is_module_available("pyrovelocity"):
        try:
            import pyrovelocity
        except (ImportError, AssertionError):
            print("pyrovelocity is not successfully installed")
            sys.exit()
    else:
        print("Installing pyrovelocity...")
        install_package("pyrovelocity==0.2.0b12")
        try:
            import pyrovelocity

@@

import pyrovelocity
print(pyrovelocity.__file__)
print(pyrovelocity.__version__)
```

@@

import pyro
print(pyro.__file__)
print(pyro.__version__)
```

@@

```{python}
#| label: print-pyrovelocity-docstring
help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
```

@@

### Data set review 
```{python}
#| label: import-ipython-display
#| echo: false
from IPython.display import Image, display
```

@@

<!-- ![Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms.](data/processed/pancreas_thresh_histogram.pdf.png){#fig-raw-counts} -->
```{python}
#| label: fig-raw-counts
#| echo: false
#| fig-cap: "Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms."
display(Image(filename=f"data/processed/pancreas_thresh_histogram.pdf.png"))
```

@@

<!-- ![Vector field](reports/pancreas_model2/vector_field.pdf.png){#fig-vector-field} -->
```{python}
#| label: fig-vector-field
#| echo: false
#| fig-cap: "Vector field" 
display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
```

@@

<!-- ![Vector fields with uncertainty estimates](reports/pancreas_model2/fig2_part1_plot.pdf.png){#fig-vector-fields-uncertainty} -->
```{python}
#| label: fig-vector-fields-uncertainty
#| echo: false
#| fig-cap: "Vector fields with uncertainty estimates."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
```

@@

<!-- ![RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene](reports/pancreas_model2/volcano.pdf.png){#fig-volcano} -->
```{python}
#| label: fig-volcano
#| echo: false
#| fig-cap: "RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/volcano.pdf.png"))
```

@@

<!-- ![Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes](reports/pancreas_model2/rainbow.pdf.png){#fig-rainbow} -->
```{python}
#| label: fig-rainbow
#| echo: false
#| fig-cap: "Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/rainbow.pdf.png"))
```

@@

<!-- ![Summary of gene selection results](reports/pancreas_model2/fig2_part2_plot.pdf.png){#fig-gene-selection-summary} -->
```{python}
#| label: fig-gene-selection-summary
#| echo: false
#| fig-cap: "Summary of gene selection results."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part2_plot.pdf.png"))
```

@@

<!-- ![Centered parameter uncertainties for a subset of genes](reports/pancreas_model2/parameter_uncertainties.pdf.png){#fig-parameter-uncertainty} -->
```{python}
#| label: fig-parameter-uncertainty
#| echo: false
#| fig-cap: "Centered parameter uncertainties for a subset of genes."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/parameter_uncertainties.pdf.png"))
```

<!-- 
TODO: save png copy of shared time plot
#### Shared time

@fig-shared-time-uncertainty shows the uncertainty in the shared time estimates for each gene.

![Uncertainty in the shared time estimates for each gene](reports/pancreas_model2/shared_time.pdf){#fig-shared-time-uncertainty} 
-->

>>>>

### Overview

Installation should take less than **5 minutes**. 
It involves checking if the notebook is running in colab, in which case it is certain that you will need to install or reinstall pyrovelocity.
Otherwise, it is assumed that the user will have installed pyrovelocity in the same environment in which the jupyter kernel is running.

@@

### Install library

@@

def setup_pyrovelocity():
    if is_module_available("pyrovelocity"):
        try:
            import pyrovelocity
        except (ImportError, AssertionError):
            print("pyrovelocity is not successfully installed")
            sys.exit()
    else:
        print("Installing pyrovelocity...")
        install_package("pyrovelocity[workflows] @ git+https://github.com/pinellolab/pyrovelocity.git@530-nbwf")
        try:
            import pyrovelocity

@@

import pyrovelocity
if IN_COLAB:
  print(pyrovelocity.__file__)
print(pyrovelocity.__version__)
```

@@

import pyro
if IN_COLAB:
  print(pyro.__file__)
print(pyro.__version__)
```

@@

```{python}
#| label: print-pyrovelocity-docstring-1
pyrovelocity.help.internal_help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
```

@@

### Data set review 
```{python}
#| label: import-ipython-display
#| echo: true
from IPython.display import Image, display
```

@@

```{python}
#| label: fig-raw-counts
#| echo: true
#| fig-cap: "Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms."

display(Image(filename=f"data/processed/pancreas_thresh_histogram.pdf.png"))
```

@@

```{python}
#| label: fig-vector-field
#| echo: true
#| fig-cap: "Vector field"

display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
```

@@

```{python}
#| label: fig-vector-fields-uncertainty
#| echo: true
#| fig-cap: "Vector fields with uncertainty estimates."

display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
```

@@

```{python}
#| label: fig-volcano
#| echo: true
#| fig-cap: "RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene."

display(Image(filename=f"{dataset_summary.data_model_reports.path}/volcano.pdf.png"))
```

@@

```{python}
#| label: fig-rainbow
#| echo: true
#| fig-cap: "Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes."

display(Image(filename=f"{dataset_summary.data_model_reports.path}/rainbow.pdf.png"))
```

@@

```{python}
#| label: fig-gene-selection-summary
#| echo: true
#| fig-cap: "Summary of gene selection results."

display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part2_plot.pdf.png"))
```

@@

```{python}
#| label: fig-parameter-uncertainty
#| echo: true
#| fig-cap: "Centered parameter uncertainties for a subset of genes."

display(Image(filename=f"{dataset_summary.data_model_reports.path}/parameter_uncertainties.pdf.png"))
```

#### Shared time

@fig-shared-time-uncertainty shows the distribution of coefficients of variation together with the mean and standard deviation (uncertainty) of the shared time for each cell. A kernel density estimate in the shared time uncertainty plot highlights the 90th percentile of the shared time standard deviation.

```{python}
#| label: fig-shared-time-uncertainty
#| echo: true
#| fig-cap: "Distribution of shared time coefficients of variation together with standard deviation (uncertainty) and mean per cell."

display(Image(filename=f"{dataset_summary.data_model_reports.path}/shared_time.pdf.png"))
```

## Experimentation

After having reviewed results that are currently computed systematically, we can also load the data and experiment.

Here we will load the data and posterior samples, perform the analysis used to generate the volcano plot for gene selection, and then regenerate the plot of phase portraits, temporal evolution, and shared time correlation. It would be straightforward, and perhaps preferable in some cases, to extract all the relevant parameters from the configuration dataclass object we referenced in each section above, but we will intentionally hard-code values to be concrete. Of course, feel free to generalize these as might suit your use case.

To load the posterior samples and postprocessed data

```{python}
#| label: load-posterior-samples
import pyrovelocity as pv
import scanpy as sc

logger = pv.logging.configure_logging("pyrovelocity.user_notebook")

posterior_samples = pv.io.CompressedPickle.load("models/pancreas_model2/pyrovelocity.pkl.zst")
adata = sc.read("models/pancreas_model2/postprocessed.h5ad")
```

to generate the marker genes

```{python}
#| label: generate-marker-genes
volcano_data = posterior_samples["gene_ranking"] 
putative_marker_genes = pv.analyze.pareto_frontier_genes(volcano_data=volcano_data, num_genes=6)
```

In @fig-manually-generated-rainbowplot we show the manually generated marker gene phase portraits, temporal evolution, and shared time correlation.

```{python}
#| label: fig-manually-generated-rainbowplot
#| echo: true
#| fig-cap: "Manually generated rainbow plot." 
with SuppressOutput():
  pv.plots.rainbowplot(
      volcano_data=volcano_data,
      adata=adata,
      posterior_samples=posterior_samples,
      genes=putative_marker_genes,
      data=["st", "ut"],
      basis="umap",
      cell_state="clusters",
  )
```

The best place to review the approach to generating the above plots is in [the source code for the `pyrovelocity.summarize` module](https://github.com/pinellolab/pyrovelocity/tree/beta/src/pyrovelocity/summarize.py) or in the [docs](https://github.io/pinellolab/pyrovelocity). Additional links are shown in the help below.

```{python}
#| label: print-pyrovelocity-docstring-2
pv.help.internal_help(pv)
```
